Short-time simulations of open collective quantum spin systems

### 2019-04-17 soc_squeezing/collective_paper/collective_spin.tex
changed order_cap in two-time correlator calculation to 20; renamed eq:TST in collective spin paper added a few missing hats on operators

We can formally expand Heisenberg operators $\hat\O\p{t}$ in a Taylor
series about the time $t=0$ to write
\begin{align}
  \hat\O\p{t}
  = e^{t\check T} \hat \O\p{0}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{\v m, \v n} \hat\S_{\v m} T^k_{\v m\v n} \O_{\v n}\p{0},
  \label{eq:time_series}
\end{align}
where the matrix elements $T^k_{\v m\v n}$ of the $k$-th time
derivative operator are
\begin{align}
  T^0_{\v m\v n} &\equiv \delta_{\v m\v n}, \\
  T^1_{\v m\v n} &\equiv T_{\v m\v n}, \\
  T^{k>1}_{\v m\v n}
  &\equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_{k-1}} \cdots T_{\v p_3\v p_2}
  T_{\v p_2\v p_1} T_{\v p_1\v n},
\end{align}
with $\delta_{\v m\v n}=1$ if $\v m=\v n$ and 0 otherwise.  For
sufficiently short times, we can truncate the series in
Eq.~\eqref{eq:time_series} by taking
\begin{align}
  \hat\O\p{t}
  \to \sum_{k=0}^M \f{t^k}{k!}
  \sum_{\v m, \v n} \hat\S_{\v m} T^k_{\v m\v n} \O_{\v n}\p{0}.
  \label{eq:order_truncation}
\end{align}
We refer Eq.~\eqref{eq:order_truncation} as the truncated short-time
(TST) expansion of Heisenberg operators.  Unlike the truncation in
Eq.~\eqref{eq:weight_truncation}, the nonzero matrix elements
$T^k_{\v m\v n}$ for $k=0,1,\cdots,M$ in
Eq.~\eqref{eq:order_truncation} tell us which operators are relevant
for computing the expectation value $\bk{\hat\O\p{t}}$ to a fixed
order $M$.  Note that using the relation
$\hat\S_{\v m}^\dag=\hat\S_{\v m^*}$, which by Hermitian conjugation
of Eq.~\eqref{eq:EOM} also implies that
$T_{\v m^*\v n^*}=T_{\v m\v n}^*$, can cut both the number of
initial-time expectation values $\bk{\hat\S_{\v m}}_{t=0}$ and the
number qof matrix elements $T_{\v m\v n}$ that we may need to
explicitly compute roughly in half.

In principle, the factorial suppression of terms at higher orders of
the expansion in Eq.~\eqref{eq:time_series} implies for any given time
$t$, there exists a truncation order $M_t$ for which the truncation
error in Eq.~\eqref{eq:order_truncation} is negligibly small.  In
practice, only a maximal truncation order $M_{\t{max}}$ is accessible
with limited computational resources, such that the TST expansion in
Eq.~\eqref{eq:order_truncation} only allows for computation of the
expectation value $\bk{\hat\O\p{t}}$ to a maximal time
$t_{M_{\t{max}}}$.  As we will see in the following section,
$M_{\t{max}}=35$ will suffice for the computation of collective spin
correlators up to times that are relevant e.g.~for spin squeezing
protocols.

@@

The Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} squeeze the
initial product state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
with $\hat S_\x\ket{\X}=S\ket{\X}$.  Our measure of spin squeezing is
the directionally-unbiased Ramsey squeezing parameter determined by
the maximal gain in resolution $\Delta\phi$ of a phase angle $\phi$
over that achieved by any spin-polarized product state
(e.g.~$\ket{\X}$)\cite{wineland1992spin, ma2011quantum},
\begin{align}
  \xi^2
  \equiv \f{\p{\Delta\phi_{\t{min}}}^2}{\p{\Delta\phi_{\t{polarized}}}^2}
  = \f{N}{\lvert\braket{\uv S}\rvert^2}
  \min_{\uv n\perp\bk{\uv S}} \Braket{\p{\uv S\cdot\uv n}^2},
  \label{eq:squeezing}
\end{align}
where $\uv S\equiv\p{\hat S_\x,\hat S_\y,\hat S_\z}$ is a collective
spin operator-valued vector, and the minimization is performed over
all unit vectors $\uv n$ in the plane orthogonal to the mean spin
vector $\bk{\uv S}$.  This squeezing parameter is entirely determined
by one- and two-spin correlators of the form $\bk{\hat S_\a}$ and
$\bk{\hat S_\a \hat S_\b}$.  For the unitary dynamics discussed in this
work, these correlators are obtainable via exact simulations of
quantum dynamics in the $\p{N+1}$-dimensional Dicke manifold of states
$\set{\ket{S,m}}$ with net spin $S$ and spin projection $m$ onto the
$z$ axis, i.e.~with $\bk{S,m|\uv S^2|S,m}=S\p{S+1}$ and
$\bk{S,m|\hat S_\z|S,m}=m$ for $m\in\set{-S,-S+1,\cdots,S}$.  In the
presence of single-spin or collective decoherence, meanwhile, these
correlators are obtainable with the collective-spin quantum trajectory
Monte Carlo method developed in ref.~\cite{zhang2018montecarlo}.  In
this work, these exact and quantum trajectory simulations will be used
to benchmark the TST expansion in Eq.~\eqref{eq:order_truncation}.

\begin{figure}
  \centering
  \subfloat[Squeezing with unitary dynamics
  \label{fig:squeezing_coherent}]
  {\includegraphics{coherent.pdf}} \\
  \subfloat[Squeezing with decoherence:
  $\gamma_-=\gamma_+=\gamma_\z=\chi$
  \label{fig:squeezing_incoherent}]
  {\includegraphics{decoherence_weak.pdf}}
  \caption{Spin squeezing of $N=10^4$ spins initially in $\ket\X$
    under \sref{fig:squeezing_coherent} unitary and
    \sref{fig:squeezing_incoherent} non-unitary dynamics, computed
    using exact methods (solid lines), quantum trajectory simulations
    (dots), and the TST expansion in Eq.~\eqref{eq:order_truncation}
    with $M=35$ (darker dashed lines).  Solid circles mark the times
    at which the TST expansion gives an unphysical result with
    $\xi^2<0$.}
  \label{fig:benchmarking}
\end{figure}

Figure \ref{fig:benchmarking} compares the squeezing parameter $\xi^2$
for $N=10^4$ spins initially in the state $\ket{\X}$ evolved under the
Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}, as computed via
both benchmarking simulations and the TST expansion in
Eq.~\eqref{eq:order_truncation} with $M=35$.  Squeezing is shown for
both unitary dynamics (Figure \ref{fig:squeezing_coherent}), as well
as non-unitary dynamics in the presence of spontaneous decay,
excitation, and dephasing of individual spins at rates $\chi$ (Figure
\ref{fig:squeezing_incoherent}), respectively described by the sets of
jump operators $\A_\a=\set{\hat s_\a^{(j)}}$ with corresponding
decoherence rates $\gamma_\a=\chi$ for $\a\in\set{-,+,\z}$.  The
results shown in Figure \ref{fig:benchmarking} were computed in a
rotated basis with
$\p{\hat s_\z,\hat s_\x}\to\p{\hat s_\x,-\hat s_\z}$ and
$\ket\X\to\ket{-\Z}\equiv\ket\dn^{\otimes N}$ (together with
appropriate transformations of the Hamiltonian and jump operators);
the only effect of this rotation on the results presented in Figure
\ref{fig:benchmarking} is to prolong the time for which the TST
expansion agrees with the benchmarking simulations.  The reason for
different results in rotated basis has to do with the breakdown of the
TST expansion, which we discuss further below.

@@

\begin{figure}
  \centering
  \includegraphics{decoherence_strong.pdf}
  \caption{Spin squeezing of $N=10^4$ spins initially in $\ket\X$ with
    spontaneous decay, excitation, and dephasing of individual spins
    at rates $\gamma_-=\gamma_+=\gamma_\z=100\chi$.  Computed using
    the TST expansion in Eq.~\eqref{eq:order_truncation} with $M=35$.
    Solid circles mark the times at which the TST expansion gives an
    unphysical result with $\xi^2<0$.}
  \label{fig:decoherence_strong}
\end{figure}

@@

\begin{figure*}
  \centering
  \includegraphics{two_time.pdf}
  \caption{Magnitude and phase of the two-time connected correlator
    $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$ defined in
    Eq.~\eqref{eq:two_time} for $N=10^4$ spins initially in the
    polarized state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
    evolving under the squeezing Hamiltonians in
    Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}.  Results are shown for time
    evolution without decoherence (leftmost panel), as well as with
    decoherence at rates $\Gamma_\a$ or $\gamma_\a$, respectively
    corresponding to collective or single-spin decoherence through
    jump operators $S_\a$ or $s_\a^{(j)}$.  Collective decoherence
    rates ($\Gamma_\a$) were chosen to be comparable to the OAT
    squeezing strength $\chi$, while single-spin decoherence rates
    ($\gamma_\a$) were scaled by an additional factor of $S=N/2$ in
    order to yield comparable effects on the correlator $C\p{t}$ at
    the times shown.  At short times, the correlations captured by
    $C\p{t}$ are essentially insensitive to collective dephasing
    ($\Gamma_\z$), while the magnitude $\abs{C\p{t}}$ is suppressed by
    other forms of decoherence.  The type of decoherence (namely spin
    decay, excitation, or dephasing) leaves a signature on the phase
    $\phi\p{t}$ of $C\p{t}$.}
  \label{fig:two_time}
\end{figure*}

>>>>

We can formally expand Heisenberg operators $\hat\O\p{t}$ in a Taylor
series about the time $t=0$ to write
\begin{align}
  \hat\O\p{t}
  = e^{t\check T} \hat \O\p{0}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{\v m, \v n} \hat\S_{\v m} T^k_{\v m\v n} \O_{\v n}\p{0},
  \label{eq:time_series}
\end{align}
where the matrix elements $T^k_{\v m\v n}$ of the $k$-th time
derivative operator are
\begin{align}
  T^0_{\v m\v n} &\equiv \delta_{\v m\v n}, \\
  T^1_{\v m\v n} &\equiv T_{\v m\v n}, \\
  T^{k>1}_{\v m\v n}
  &\equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_{k-1}} \cdots T_{\v p_3\v p_2}
  T_{\v p_2\v p_1} T_{\v p_1\v n},
\end{align}
with $\delta_{\v m\v n}=1$ if $\v m=\v n$ and 0 otherwise.  For
sufficiently short times, we can truncate the series in
Eq.~\eqref{eq:time_series} by taking
\begin{align}
  \hat\O\p{t}
  \to \sum_{k=0}^M \f{t^k}{k!}
  \sum_{\v m, \v n} \hat\S_{\v m} T^k_{\v m\v n} \O_{\v n}\p{0}.
  \label{eq:TST}
\end{align}
We refer Eq.~\eqref{eq:TST} as the truncated short-time (TST)
expansion of Heisenberg operators.  Unlike the truncation in
Eq.~\eqref{eq:weight_truncation}, the nonzero matrix elements
$T^k_{\v m\v n}$ for $k=0,1,\cdots,M$ in Eq.~\eqref{eq:TST} tell us
which operators are relevant for computing the expectation value
$\bk{\hat\O\p{t}}$ to a fixed order $M$.  Note that using the relation
$\hat\S_{\v m}^\dag=\hat\S_{\v m^*}$, which by Hermitian conjugation
of Eq.~\eqref{eq:EOM} also implies that
$T_{\v m^*\v n^*}=T_{\v m\v n}^*$, can cut both the number of
initial-time expectation values $\bk{\hat\S_{\v m}}_{t=0}$ and the
number qof matrix elements $T_{\v m\v n}$ that we may need to
explicitly compute roughly in half.

In principle, the factorial suppression of terms at higher orders of
the expansion in Eq.~\eqref{eq:time_series} implies for any given time
$t$, there exists a truncation order $M_t$ for which the truncation
error in Eq.~\eqref{eq:TST} is negligibly small.  In practice, only a
maximal truncation order $M_{\t{max}}$ is accessible with limited
computational resources, such that the TST expansion in
Eq.~\eqref{eq:TST} only allows for computation of the expectation
value $\bk{\hat\O\p{t}}$ to a maximal time $t_{M_{\t{max}}}$.  As we
will see in the following section, $M_{\t{max}}=35$ will suffice for
the computation of collective spin correlators up to times that are
relevant e.g.~for spin squeezing protocols.

@@

The Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} squeeze the
initial product state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
with $\hat S_\x\ket{\X}=S\ket{\X}$.  Our measure of spin squeezing is
the directionally-unbiased Ramsey squeezing parameter determined by
the maximal gain in resolution $\Delta\phi$ of a phase angle $\phi$
over that achieved by any spin-polarized product state
(e.g.~$\ket{\X}$)\cite{wineland1992spin, ma2011quantum},
\begin{align}
  \xi^2
  \equiv \f{\p{\Delta\phi_{\t{min}}}^2}{\p{\Delta\phi_{\t{polarized}}}^2}
  = \f{N}{\lvert\braket{\uv S}\rvert^2}
  \min_{\uv n\perp\bk{\uv S}} \Braket{\p{\uv S\cdot\uv n}^2},
  \label{eq:squeezing}
\end{align}
where $\uv S\equiv\p{\hat S_\x,\hat S_\y,\hat S_\z}$ is a collective
spin operator-valued vector, and the minimization is performed over
all unit vectors $\uv n$ in the plane orthogonal to the mean spin
vector $\bk{\uv S}$.  This squeezing parameter is entirely determined
by one- and two-spin correlators of the form $\bk{\hat S_\a}$ and
$\bk{\hat S_\a \hat S_\b}$.  For the unitary dynamics discussed in
this work, these correlators are obtainable via exact simulations of
quantum dynamics in the $\p{N+1}$-dimensional Dicke manifold of states
$\set{\ket{S,m}}$ with net spin $S$ and spin projection $m$ onto the
$z$ axis, i.e.~with $\bk{S,m|\uv S^2|S,m}=S\p{S+1}$ and
$\bk{S,m|\hat S_\z|S,m}=m$ for $m\in\set{-S,-S+1,\cdots,S}$.  In the
presence of single-spin or collective decoherence, meanwhile, these
correlators are obtainable with the collective-spin quantum trajectory
Monte Carlo method developed in ref.~\cite{zhang2018montecarlo}.  In
this work, these exact and quantum trajectory simulations will be used
to benchmark the TST expansion in Eq.~\eqref{eq:TST}.

\begin{figure}
  \centering
  \subfloat[Squeezing with unitary dynamics
  \label{fig:squeezing_coherent}]
  {\includegraphics{coherent.pdf}} \\
  \subfloat[Squeezing with decoherence:
  $\gamma_-=\gamma_+=\gamma_\z=\chi$
  \label{fig:squeezing_incoherent}]
  {\includegraphics{decoherence_weak.pdf}}
  \caption{Spin squeezing of $N=10^4$ spins initially in $\ket\X$
    under \sref{fig:squeezing_coherent} unitary and
    \sref{fig:squeezing_incoherent} non-unitary dynamics, computed
    using exact methods (solid lines), quantum trajectory simulations
    (dots), and the TST expansion in Eq.~\eqref{eq:TST} with $M=35$
    (darker dashed lines).  Solid circles mark the times at which the
    TST expansion gives an unphysical result with $\xi^2<0$.}
  \label{fig:benchmarking}
\end{figure}

Figure \ref{fig:benchmarking} compares the squeezing parameter $\xi^2$
for $N=10^4$ spins initially in the state $\ket{\X}$ evolved under the
Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}, as computed via
both benchmarking simulations and the TST expansion in
Eq.~\eqref{eq:TST} with $M=35$.  Squeezing is shown for both unitary
dynamics (Figure \ref{fig:squeezing_coherent}), as well as non-unitary
dynamics in the presence of spontaneous decay, excitation, and
dephasing of individual spins at rates $\chi$ (Figure
\ref{fig:squeezing_incoherent}), respectively described by the sets of
jump operators $\A_\a\equiv\set{\hat s_\a^{(j)}}$ with corresponding
decoherence rates $\gamma_\a=\chi$ for $\a\in\set{-,+,\z}$.  The
results shown in Figure \ref{fig:benchmarking} were computed in a
rotated basis with
$\p{\hat s_\z,\hat s_\x}\to\p{\hat s_\x,-\hat s_\z}$ and
$\ket\X\to\ket{-\Z}\equiv\ket\dn^{\otimes N}$ (together with
appropriate transformations of the Hamiltonian and jump operators);
the only effect of this rotation on the results presented in Figure
\ref{fig:benchmarking} is to prolong the time for which the TST
expansion agrees with the benchmarking simulations.  The reason for
different results in rotated basis has to do with the breakdown of the
TST expansion, which we discuss further below.

@@

\begin{figure}
  \centering
  \includegraphics{decoherence_strong.pdf}
  \caption{Spin squeezing of $N=10^4$ spins initially in $\ket\X$ with
    spontaneous decay, excitation, and dephasing of individual spins
    at rates $\gamma_-=\gamma_+=\gamma_\z=100\chi$.  Computed using
    the TST expansion in Eq.~\eqref{eq:TST} with $M=35$.  Solid
    circles mark the times at which the TST expansion gives an
    unphysical result with $\xi^2<0$.}
  \label{fig:decoherence_strong}
\end{figure}

@@

\begin{figure*}
  \centering
  \includegraphics{two_time.pdf}
  \caption{Magnitude and phase of the two-time connected correlator
    $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$ defined in
    Eq.~\eqref{eq:two_time} for $N=10^4$ spins initially in the
    polarized state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
    evolving under the squeezing Hamiltonians in
    Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}.  Results are shown for time
    evolution without decoherence (leftmost panel), as well as with
    decoherence at rates $\Gamma_\a$ or $\gamma_\a$, respectively
    corresponding to collective or single-spin decoherence through
    jump operators $\hat S_\a$ or $\hat s_\a^{(j)}$.  Collective
    decoherence rates ($\Gamma_\a$) were chosen to be comparable to
    the OAT squeezing strength $\chi$, while single-spin decoherence
    rates ($\gamma_\a$) were scaled by an additional factor of $S=N/2$
    in order to yield comparable effects on the correlator $C\p{t}$ at
    the times shown.  At short times, the correlations captured by
    $C\p{t}$ are essentially insensitive to collective dephasing
    ($\Gamma_\z$), while the magnitude $\abs{C\p{t}}$ is generally
    suppressed by other forms of decoherence.  The general type of
    decoherence (namely spin decay, excitation, or dephasing) also
    leaves a signature on the phase $\phi\p{t}$ of $C\p{t}$.  Results
    computed using the TST expansion in Eq.~\eqref{eq:TST} with
    $M=20$.}
  \label{fig:two_time}
\end{figure*}
